---
always_allow_html: yes
output:
  html_document: default
  tufte::tufte_handout: default
---

## nextHIV Report {.tabset}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r loadlibraries}
suppressPackageStartupMessages(library(rethinker))
suppressPackageStartupMessages(library(ape))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggforce))
suppressPackageStartupMessages(library(ggtree))
suppressPackageStartupMessages(library(epiwidgets))
suppressPackageStartupMessages(library(cowplot))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(DT))
suppressPackageStartupMessages(library(xtable))
suppressPackageStartupMessages(library(gamlss))
suppressPackageStartupMessages(library(gamlss.dist))
suppressPackageStartupMessages(library(gamlss.mx))
```

```{r setglobals}
host <- "localhost"
port <- 28015
authKey <- NULL
v <- "V0_4"
db <- "nexthiv"
pid <- "CFAR_PID"
```

```{r definefunctions}
get_list <- function(db,tbl,maxResults=10000,host="localhost",port=28015,authKey=NULL,v="V0_4"){
  connection <- openConnection(host = host, port = port, authKey = authKey, v = v)
  r() -> query
  query$db(db)
  query$table(tbl)
  cursor <- query$run(connection)
  l <- cursorToList(cursor,maxResults=maxResults)
  close(connection)
  return(l)
}
get_table <- function(db,tbl,maxResults=10000,host="localhost",port=28015,authKey=NULL,v="V0_4"){
  l <- get_list(db,tbl,maxResults,host,port,authKey,v)
  df <- do.call(rbind, lapply(l, data.frame, stringsAsFactors=FALSE))
  return(df)
}
list_to_df <- function(l){
  df <- do.call(rbind, lapply(l, data.frame, stringsAsFactors=FALSE))
  return(df)
}
extract_annotations <- function(a,col,sep){
  strsplit(a,sep) %>% lapply(.,"[",col) %>% unlist
}
txt_to_dna <- function(nm,s){
y <- t(sapply(strsplit(s,""), tolower))
rownames(y) <- nm
as.DNAbin(y)
}
```


```{r definethemes}
blank_theme <- theme_minimal()+
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold")
  )
```


```{r loadtables}
demographics <- get_table(db,"demographics")
sequences <- get_table(db,"sequences")
regimens <- get_table(db,"regimens")
processed_sequences <- get_table(db,"processed_sequences")
clustering <- get_table(db,"clustering")
resistance <- get_list(db,"resistance") %>% lapply(.,"[",c("id","BASELINE","PI_MUTATION_COUNT","NRTI_MUTATION_COUNT","NNRTI_MUTATION_COUNT")) %>% list_to_df
#distances <- get_table(db,"distances")
```

```{r processregimens}
# naive
o <- order(regimens$CFAR_PID,regimens$AGE_AT_START)
regimens2 <- regimens[o,]
regimens2 <- regimens2[!duplicated(regimens2$CFAR_PID),]
```



```{r makemastertable}
#Note that as clustering only has baseline sequences, the following only has 2914 individuals
master <- merge(sequences,clustering,by.x="id",by.y="id")
master <- merge(master,processed_sequences,by.x="id",by.y="id")
master <- merge(master,resistance,by.x="id",by.y="id")
master <- merge(master,demographics,by="CFAR_PID")
master <- merge(master,regimens2,by="CFAR_PID",all.x=TRUE,all.y=FALSE)
o <- order(master$CFAR_PID,as.double(master$AGE_AT_SEQUENCE))
master <- master[o,]
baseline <- master#[master$BASELINE==TRUE,]
```

```{r identifynaive}
naive <- baseline$AGE_AT_SEQUENCE<baseline$AGE_AT_START
naive[is.na(naive)] <- TRUE
baseline$naive <- naive
```


```{r recentdiagnosis,eval=FALSE}
recentd <- baseline$AGE_AT_ENROLLMENT-baseline$AGE_AT_HIV_DX<1./12
recentd[is.na(recentd)] <- TRUE
baseline$recentd <- recentd
```


```{r,eval=FALSE}
write.table(demographics,file="demographics.txt",col.names=TRUE,row.names=FALSE,sep="\t",quote=FALSE)
write.table(baseline,file="baseline.txt",col.names=TRUE,row.names=FALSE,sep="\t",quote=FALSE)
```

### Summary

Date of report: `r Sys.Date()`

No. of individuals: `r length(unique(sequences$CFAR_PID))`

No. of sequences: `r length(sequences$CFAR_PID)`

### Demographics


```{r}
p <- ggplot(demographics,aes(x=as.double(YEAR_OF_COHORT_ENTRY)))+geom_bar()+xlab("Year of cohort entry")+ylab("Count")
ggplotly(p)
```


```{r}
p <- ggplot(demographics,aes(x=as.double(AGE_AT_ENROLLMENT)))+geom_histogram(binwidth=1)+xlab("Age at cohort entry")+ylab("Count")
ggplotly(p)
```


```{r}
p <- ggplot(demographics,aes(x=SEX))+geom_bar()+xlab("Gender")+ylab("Count")+scale_x_discrete(labels=c("Female","Male"))
ggplotly(p)
```

```{r}
p <- ggplot(demographics,aes(x=RISK1))+geom_bar()+xlab("Risk behavior")+ylab("Count")
ggplotly(p)
```


```{r}
zip3 <- demographics$postCode3
ztbl <- table(zip3)
rarezip <- names(ztbl)[ztbl<=5]
zip3[zip3%in%rarezip] <- "Other"
x <- data.frame(ZIP3=zip3)
p <- ggplot(x,aes(x=ZIP3))+geom_bar()+xlab("ZIP3 code")+ylab("Count")
ggplotly(p)
```


```{r}
cntry <- demographics$COUNTRY_OF_ORIGIN
cntry[cntry==""] <- "MISSING"
ctbl <- table(cntry)
rarecntry <- names(ctbl)[ctbl<=5]
cntry[cntry%in%rarecntry] <- "OTHER"
x <- data.frame(Country=factor(cntry,levels=names(sort(table(cntry),decreasing=TRUE))))
p <- ggplot(x,aes(x=Country))+geom_bar()+xlab("Country of origin")+ylab("Count")+ theme(axis.text.x = element_text(angle = 90, hjust = -1))
ggplotly(p)
```

```{r racecount}
races <- c("Undefined","Caucasian","African American","Hispanic","Native American","Asian","Pacific Islander","Indian","Middle Eastern","African","Mixed")
race <- as.integer(demographics$RACE)
race <- race+1
race <- races[race]
rtbl <- table(race)
rarerace <- names(rtbl)[rtbl<=5]
race[race%in%rarerace] <- "Other"
x <- data.frame(Race=factor(race,levels=names(sort(table(race),decreasing=TRUE))))
p <- ggplot(x,aes(x=Race))+geom_bar()+xlab("Race")+ylab("Count")+ theme(axis.text.x = element_text(angle = 90, hjust = -1))
ggplotly(p)
```

### Sequences


```{r seqperperson}
df <- as.data.frame.table(table(table(sequences$CFAR_PID)))
names(df) <- c("Number","Count")
p <- ggplot(df,aes(x=Number,y=Count))+geom_bar(stat="identity")+xlab("Sequences per individual")+ylab("Count")
ggplotly(p)
```

<br>

```{r}
datatable(df)
```

### Resistance

#### PI resistance mutations

Naive:

```{r}
df <- as.data.frame.table(table(baseline$PI_MUTATION_COUNT[baseline$naive==TRUE]))
names(df) <- c("Number","Count")
p <- ggplot(df,aes(x="",y=Count,fill=Number))+geom_bar(width=1,stat="identity")+coord_polar("y", start=0)+blank_theme + theme(axis.text.x=element_blank())
p
```

Experienced:

```{r}
df <- as.data.frame.table(table(baseline$PI_MUTATION_COUNT[baseline$naive==FALSE]))
names(df) <- c("Number","Count")
p <- ggplot(df,aes(x="",y=Count,fill=Number))+geom_bar(width=1,stat="identity")+coord_polar("y", start=0)+blank_theme + theme(axis.text.x=element_blank())
p
```
#### NRTI resistance mutations

Naive:

```{r nrtiresistancenaive}
df <- as.data.frame.table(table(baseline$NRTI_MUTATION_COUNT[baseline$naive==TRUE]))
names(df) <- c("Number","Count")
p <- ggplot(df,aes(x="",y=Count,fill=Number))+geom_bar(width=1,stat="identity")+coord_polar("y", start=0)+blank_theme + theme(axis.text.x=element_blank())
p
```

Experienced:

```{r nrtiresistanceexp}
df <- as.data.frame.table(table(baseline$NRTI_MUTATION_COUNT[baseline$naive==FALSE]))
names(df) <- c("Number","Count")
p <- ggplot(df,aes(x="",y=Count,fill=Number))+geom_bar(width=1,stat="identity")+coord_polar("y", start=0)+blank_theme + theme(axis.text.x=element_blank())
p
```

#### NNRTI resistance mutations

Naive:

```{r nnrtiresistancenaive}
df <- as.data.frame.table(table(baseline$NNRTI_MUTATION_COUNT[baseline$naive==TRUE]))
names(df) <- c("Number","Count")
p <- ggplot(df,aes(x="",y=Count,fill=Number))+geom_bar(width=1,stat="identity")+coord_polar("y", start=0)+blank_theme + theme(axis.text.x=element_blank())
p
```

Experienced:

```{r nnrtiresistanceexp}
df <- as.data.frame.table(table(baseline$NNRTI_MUTATION_COUNT[baseline$naive==FALSE]))
names(df) <- c("Number","Count")
p <- ggplot(df,aes(x="",y=Count,fill=Number))+geom_bar(width=1,stat="identity")+coord_polar("y", start=0)+blank_theme + theme(axis.text.x=element_blank())
p
```


### Subtypes

#### Breakdown

```{r}
subtypes <- extract_annotations(baseline$REFALTER,2,"\\|")
df <- as.data.frame.table(table(subtypes))
names(df) <- c("Subtype","Count")
datatable(df)
```


#### Non B subtypes

```{r nonbsubtypes}
df2 <- df[df$Subtype!="B",]
o <- order(df2$Count,decreasing=TRUE)
df2 <- df2[o,]
df2$Subtype <- factor(df2$Subtype,levels=df2$Subtype)
p <- ggplot(df2,aes(x=Subtype,y=Count,fill=Subtype))+geom_bar(width=1,stat="identity")+scale_y_continuous()+theme(axis.text.x = element_text (angle = 90, vjust = 0.5, size = 16),legend.position="none")
ggplotly(p)
```

### Clustering

```{r,eval=FALSE}
thresh <- 0.001
vars <- c("RISK1","SEX","RACE","AGE_AT_ENROLLMENT")
varlabels <- c("Risk","Sex","Race","Age")
nvars <- length(vars)
df0 <- baseline[,c("MINDST","REFMINDST",vars)]
df0$MINDST2 <- df0$MINDST
df0$MINDST2[df0$MINDST<=thresh] <- 0.0
df0$REFMINDST2 <- df0$REFMINDST
df0$REFMINDST2[df0$REFMINDST<=thresh] <- 0.0
df0$AGE_AT_ENROLLMENT <- as.double(df0$AGE_AT_ENROLLMENT)
results <- list()
for(i in 1:nvars){
  print(varlabels[i])
  v <- vars[i]
  df <- na.omit(df0[,c("MINDST2",v)])
  f1 <- paste("MINDST2 ~ ",v,sep="")
  f2 <- paste("~",v)
  g <- gamlss(as.formula(f1),nu.formula=as.formula(f2),data=df,family=ZAGA)
  s <- summary(g)
  r <- s[grep(v,row.names(s)),]
  results[[v]] <- r
}
```

```{r,eval=FALSE}
vars <- c("RISK1","SEX","RACE","AGE_AT_ENROLLMENT")
varlabels <- c("Risk","Sex","Race","Age")
nvars <- length(vars)
df0 <- baseline[,c("MINDST","REFMINDST",vars)]
df0$MINDST[df0$MINDST<0.001] <- 0.001
df0$AGE_AT_ENROLLMENT <- as.double(df0$AGE_AT_ENROLLMENT)
results <- list()
for(i in 1:nvars){
  print(varlabels[i])
  v <- vars[i]
  df <- na.omit(df0[,c("MINDST",v)])
  f1 <- paste("MINDST ~ ",v,sep="")
  f2 <- paste("~",v)
  g <- gamlssMX(as.formula(f1),pi.formula=as.formula(f2),data=df,family="GA",K=2)
  s <- summary(g)
  r <- s[grep(v,row.names(s)),]
  results[[v]] <- r
}
```


#### Distribution of closest sequences

```{r}
d <- baseline
d <- d[d$MINDST>=0,]
o <- order(d$CFAR_PID,d$MINDST)
idx <- !duplicated(d$CFAR_PID)
d <- d[idx,]
p <- ggplot(d,aes(x=as.double(MINDST)))+geom_histogram(binwidth=0.001)+xlab("Minimum genetic distance")+ylab("Count")
ggplotly(p)
```

#### Minimum distance by gender

```{r mindstgender}
p <- ggplot(d,aes(y=as.double(MINDST),x=SEX))+geom_sina(aes(col=SEX))+xlab("Gender")+ylab("Distance")+scale_x_discrete(labels=c("Female","Male"))+theme(legend.position="none")
p
```

#### Minimum distance by risk factor

```{r mindstrisk}
p <- ggplot(d,aes(y=as.double(MINDST),x=RISK1))+geom_sina(aes(col=RISK1))+xlab("Risk behavior")+ylab("Distance")+theme(legend.position="none")
p
```

#### Minimum distance by age

```{r}
p <- ggplot(d,aes(y=as.double(MINDST),x=factor(AGE_AT_SEQUENCE>30)))+geom_sina(aes(col=factor(AGE_AT_SEQUENCE>30)))+xlab("Age")+ylab("Distance")+theme(legend.position="none")+scale_x_discrete(labels=c("<30","30+"))
p
```

#### Minimum distance by year of diagnosis

```{r}
dx <- as.double(d$YEAR_OF_HIV_DX)
dx[is.na(dx)] <- d$YEAR_OF_COHORT_ENTRY[is.na(dx)]
d$dx <- dx
```


```{r}
p <- ggplot(d,aes(y=as.double(MINDST),x=dx))+geom_sina(aes(col=dx))+xlab("Year of diagnosis")+ylab("Distance")+theme(legend.position="none")+theme(axis.text.x = element_text(angle=90,hjust=0.1))
p
```

### Phylogenies

```{r}
baseline.s <- txt_to_dna(as.character(baseline$CFAR_PID),as.character(baseline$SEQUENCE_ALIGNED))
subtypes <- extract_annotations(as.character(baseline$REFALTER),2,"\\|")
```

#### Subtype C

```{r}
s <- baseline.s[subtypes=="C",]
d <- dist.dna(s,model="TN93")
tr <- nj(d)
ggtree(tr)+theme_tree2()
```

#### Subtype BF

```{r}
s <- baseline.s[subtypes=="BF",]
d <- dist.dna(s,model="TN93")
tr <- nj(d)
ggtree(tr)+theme_tree2()
```

#### Subtype A1

```{r}
s <- baseline.s[subtypes=="A1",]
d <- dist.dna(s,model="TN93")
tr <- nj(d)
ggtree(tr)+theme_tree2()
```

#### Subtype CRF02_AG

```{r}
s <- baseline.s[subtypes=="02AG",]
d <- dist.dna(s,model="TN93")
tr <- nj(d)
ggtree(tr)+theme_tree2()
```

#### Subtype CRF01_AE

```{r}
s <- baseline.s[subtypes=="01AE",]
d <- dist.dna(s,model="TN93")
tr <- nj(d)
ggtree(tr)+theme_tree2()
```